---
title: "Creating New Geom_*() Objects for {ggplot2}"
author: "Ben Stano"
date: "3/27/2022"
output: html_document
---

# Extending `{ggplot2}`

# Imports (move?)

```{r}
library(ggplot2)
library(grid)
```

# Background

The `{ggplot2}` package is the most common used and robust plotting package in R.

While the tools provided to users are straightforward once learned, extending the functionality of `{ggplot2}` is a different beast entirely. `{ggplot2}` is built on the `ggproto` system of object oriented programming, a system used only in `{ggplot2}`. This OOP system is combined with the `grob` object-type from the `{grid}` package to create the stylish graphs most R users are familiar with.

This vignette will focus on given readers enough information of the `ggproto` OOP system and `grob` objects to extend `{ggplot2}` functionality, creating graphics customized to the user's needs. For a deeper dive into `ggproto`, `grob` objects, and other `{ggplot2}` topics, use this attached references:
 - `ggproto` reference
 - `{grid}`/`grob` reference
 - `{ggplot2}` reference
 - other references?

All geom and stat type objects in `{ggplot2}` are build on top of the `ggproto` object oriented system.

Stats are concerned with applying a particular type of transformation to a dataset, while Geoms are concerned with graphically displaying given data in a particular way. Though these may seem like very different goals, both are more similar than they appear.

Both require a `ggproto` object declaration and a call to the `layer()` function.

## Creating Stats and Geoms: the `ggproto()` Function

`ggproto` objects are initialized using the `ggproto` function. At the basic level, they require a name, a character object, and a type to inherit from, for our purposed either `Stat` or `Geom`.

```{r}
G <- ggproto("G", # the name of the ggproto object
             NULL, # the inheritance type, typically `Stat` or `Geom`
             z = 1,
             double_z = function(self) {
               self$z <- self$z * 2
             }
    )

G$z
G$double_z()
G$double_z()
G$z
```

### The Stat `ggproto`

Stats are `ggproto` objects which must be constructed with at least these two arguments:

 - `required_aes`: character vector that list the aesthetics users must provide
 - `compute_group`: a function which returns properly transformed data
 
After these two arguments, a `layer()` function is used to specify defaults and specific parameters for the stat. First, one defines the defaults for the function and then the `layer()` function is called and sends arguments for the stat to the params argument. The same process is done for geoms. However, stats are more concerned with applying specific mathematically transformative logic to data. 

### The Geom `ggproto`

Creating Geoms is slightly more complicated than creating new Stats, as Geoms require an understanding of `{grid}` and `grob` objects from `{grid}`.

Geoms are `ggproto` objects which must be constructed using at least these four arguments:

 - `required_aes`: character vector that lists the aesthetics users must provide
 - `default_aes` : list of aesthetics that have default values
 - `draw_key`: provides the function used to draw the key in the legend
 - `draw_panel()`: a function that requires three arguments a returns a grid glob type object. `draw_group()` may be used instead if one graphic element per row is required. Both require the same three arguments:
   - `data`: data.frame with one column for each aesthetic
   - `panel_params`: list of per-panel parameters generated by `coord()`
   - `coord`: an object describing a coordinate system

Geoms are concerned with graphing data in a particular way.

### The `layer()` function

The layer function, regardless of whether one is creating a stat or a geom, is where the defaults from the function are sent to when calling the function. They specify arguments for the geom/stat as well as any aesthetics. The params argument is perhaps one of the most important, as it can be used to inherit parameters from other functions. \n

Essential Arguments (these are the arguments we used in the vignette):

 - `geom`: geometric object used in displaying data
 - `stat`: statistical transoformation used on the data
 - `data`: data taking the form of a data.frame, data inherited from plot data in the `ggplot()` call, or a function with the return value of a data.frame
 - `mapping`: mappings created by `aes()` to specify x and y values, for example
 - `position`: position adjustment
 - `show.legend`: whether this layer will be included in legends
 - `inherit.aes`: combines default aesthetics if TRUE, overrides them if FALSE
 - `params`: any additional parameters added to geom and stat, such as removing NA values
 
Other arguments include:

 - `check.aes`
 - `check.param`
 - `key_glyph`
 - `layer_class`

Note: While layer can also be called using `geom_*` or `stat_`, this may cause some confusion, as a layer necessitates both a geom and a state in its arguments

# Examples

## Stat Example

```{r}
StatPercentile <- ggproto("StatPercentile", Stat,
                          
  compute_group = function(data, scales) {
    data <- data.frame(y = data$y)
    new_data <- data.frame(y = apply(data, 2, function(x) ecdf(x)(sort(x))))
    new_data$x <- 1:nrow(new_data)
    new_data
  },
  
  required_aes = c("y")
  
)

stat_percentile <- function(mapping = NULL, data = NULL, geom = "line",
                       position = "identity", na.rm = FALSE, show.legend = NA, 
                       inherit.aes = TRUE) {
  layer(
    stat = StatPercentile, data = data, mapping = mapping, geom = geom, 
    position = position, show.legend = show.legend, inherit.aes = inherit.aes,
    params = list(na.rm = na.rm)
  )
}

ggplot(mpg, aes(y = hwy)) +
  stat_percentile()
```

## Allowing for Additional User Arguments

```{r}
StatPercentile <- ggproto("StatPercentile", Stat,
                          
  compute_group = function(data, scales) {
    data <- data.frame(y = data$y)
    new_data <- data.frame(y = apply(data, 2, function(x) ecdf(x)(sort(x))))
    new_data$x <- 1:nrow(new_data)
    new_data
  },
  
  required_aes = c("y")
  
)

stat_percentile <- function(mapping = NULL, data = NULL, geom = "line",
                       position = "identity", na.rm = FALSE, show.legend = NA, 
                       inherit.aes = TRUE, ...) {
  layer(
    stat = StatPercentile, data = data, mapping = mapping, geom = geom, 
    position = position, show.legend = show.legend, inherit.aes = inherit.aes,
    params = list(na.rm = na.rm, ...)
  )
}

ggplot(mpg, aes(y = hwy)) +
  stat_percentile(color = "red", linetype = "dashed", geom = "point")
```

## Geom Example

```{r}
GeomSimpleLine <- ggproto("GeomSimpleLine", Geom,
  required_aes = c("x", "y"),
  default_aes = aes(shape = 19, colour = "black"),
  draw_key = draw_key_path,

  draw_panel = function(data, panel_params, coord) {
    
    data <- dplyr::arrange(data, x)
    
    coords <- coord$transform(data, panel_params)
    grid::linesGrob(
      x = unit(coords$x, "npc"), 
      y = unit(coords$y, "npc"),
      gp = grid::gpar(col = coords$colour)
    )
  }
)

geom_simple_line <- function(mapping = NULL, data = NULL, stat = "identity",
                              position = "identity", na.rm = FALSE, show.legend = NA, 
                              inherit.aes = TRUE, ...) {
  layer(
    geom = GeomSimpleLine, mapping = mapping,  data = data, stat = stat, 
    position = position, show.legend = show.legend, inherit.aes = inherit.aes,
    params = list(na.rm = na.rm)
  )
}

ggplot(mtcars, aes(mpg, wt)) +
  geom_simple_line()
```

# Extending `{ggplot2}`: Difference Plot

Text 

## Custom Stat for Difference Plot

```{r}
StatDifference <- ggproto("StatDifference", Stat,
  compute_group = function(data, scales) {
    method1 <- (data$x + data$y)/2
    method2 <- data$x - data$y
    new_data <- data.frame(x = method1,
                           y = method2)
    new_data
  },
  
  required_aes = c("x", "y")
)
```

## Custom Geom for Difference Plot

```{r}
GeomSimpleDifference <- ggproto("GeomSimpleDifference", Geom,
  required_aes = c("x", "y"),
  default_aes = aes(shape = 20, color = "black"),
  draw_key = draw_key_point,
  
  draw_panel = function(data, panel_params, coord) {
    coords <- coord$transform(data, panel_params)
    gList(grid::pointsGrob(
      coords$x, coords$y,
      pch = coords$shape,
      gp = grid::gpar(col = coords$colour)
    ),
    grid::linesGrob(
      coords$x, mean(coords$y),
      gp = grid::gpar(col = coords$colour)
    ))
  }
  
)

geom_simple_diff <- function(mapping = NULL, data = NULL, stat = "difference",
                              position = "identity", na.rm = FALSE, show.legend = NA, 
                              inherit.aes = TRUE) {
  layer(
    geom = GeomSimpleDifference, mapping = mapping,  data = data, stat = stat, 
    position = position, show.legend = show.legend, inherit.aes = inherit.aes,
    params = list(na.rm = na.rm)
  )
}
```

## Plotting the Difference Plot

```{r}
set.seed(1)
data <- data.frame(x = runif(100, 0, 10),
                   y = runif(100, 0, 10))


ggplot(data, aes(x, y)) +
  geom_simple_diff()
```


# Exercises

```{r}
new_data <- data.frame(y = mtcars$mpg)
    model <- lm(mpg ~ wt, data = mtcars)
    new_data$x <- predict(model, newdata = data.frame(wt = mtcars$wt))
    new_data
```
```{r}
rng <- range(c(1,2,100))

?seq
data.frame(x = seq(rng[1], rng[2]))
```

1. Create a Stat for a Regression Line (`geom_smooth()`)



```{r, echo = FALSE}
StatLm <- ggproto("StatLm", Stat,
                          
  compute_group = function(data, scales) {
    rng <- range(data$x, na.rm = TRUE)
    new_data <- data.frame(x = seq(rng[1], rng[2], length.out = nrow(data)))
    
    model <- lm(y ~ x, data = data)
    
    new_data$y <- predict(model, newdata = data.frame(x = new_data$x))
    new_data
  },
  
  required_aes = c("x", "y")
  
)

stat_lm <- function(mapping = NULL, data = NULL, geom = "line",
                       position = "identity", na.rm = FALSE, show.legend = NA, 
                       inherit.aes = TRUE) {
  layer(
    stat = StatLm, data = data, mapping = mapping, geom = geom, 
    position = position, show.legend = show.legend, inherit.aes = inherit.aes,
    params = list(na.rm = na.rm)
  )
}
```
```{r}
ggplot(mtcars, aes(x = wt, y = mpg)) +
  geom_point() +
  stat_lm()
```


2. Add arguments to the Regression Line Stat



```{r, echo = FALSE}
StatLm <- ggproto("StatLm", Stat,
                          
  compute_group = function(data, scales, formula = y ~ x) {
    rng <- range(data$x, na.rm = TRUE)
    new_data <- data.frame(x = seq(rng[1], rng[2], length.out = nrow(data)))
    
    model <- lm(formula, data = data)
    
    new_data$y <- predict(model, newdata = data.frame(x = new_data$x))
    new_data
  },
  
  required_aes = c("x", "y")
  
)

stat_lm <- function(mapping = NULL, data = NULL, geom = "line",
                       position = "identity", na.rm = FALSE, show.legend = NA, 
                       inherit.aes = TRUE, ...) {
  layer(
    stat = StatLm, data = data, mapping = mapping, geom = geom, 
    position = position, show.legend = show.legend, inherit.aes = inherit.aes,
    params = list(na.rm = na.rm, ...)
  )
}
```

```{r}
ggplot(mtcars, aes(x = wt, y = mpg)) +
  geom_point() +
  stat_lm(formula = y ~ poly(x, 4), color = "blue", linetype = "dotted", size = 3)
```


3. Create a Geom for <<Something>>

```{r}
GeomSimpleHist <- ggproto("GeomSimpleHist", Geom,
  required_aes = c("x"),
  default_aes = aes(shape = 19, colour = "black"),
  draw_key = draw_key_path,

  draw_panel = function(data, panel_params, coord) {
    
    data <- dplyr::arrange(data, x)
    
    coords <- coord$transform(data, panel_params)
    grid::segmentsGrob(
      coords$x,
      gp = grid::gpar(col = coords$colour)
    )
  }
)

?draw_key
geom_simple_line <- function(mapping = NULL, data = NULL, stat = "identity",
                              position = "identity", na.rm = FALSE, show.legend = NA, 
                              inherit.aes = TRUE, ...) {
  layer(
    geom = GeomSimpleLine, mapping = mapping,  data = data, stat = stat, 
    position = position, show.legend = show.legend, inherit.aes = inherit.aes,
    params = list(na.rm = na.rm)
  )
}

ggplot(mtcars, aes(mpg, wt)) +
  geom_simple_line()
```











