---
title: "ggedf-vignette"
author: Kathryn Cronquist, Thea De Jong, and Ben Stano
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ggedf-vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(ggedf)
```

# The Empirical Distribution Function

This package uses an extension of the `ggplot2` package in order to create more
geoms to visualize the empirical distribution function. This function takes a vector, 
or distribution X and the function itself finds the proportions of each value of X that
is less or equal to the current value. It is most often used when creating a
step-function plot, which is shown below.

## The Step-Function Plot

```{r}
x <- c(1, 5, 10, 15)
edf_example <- ecdf(x)

plot(edf_example)
```

# Cumulative Percentage Polygons

Another way to visualize this function is via cumulative percentage polygons, 
which connect each value of X to show the cumulative value at each "step". Using 
the function `stat_cum_per_poly()`, one can create a plot that returns a ggplot2 
object displaying a cumulative percentage polygon as a line graph. The only 
argument required is the vector X, as well as any aesthetics normally used in
`ggplot2` such as color and linetype. 

```{r}
#example here - should it be the same as in the help page? 
```


# Percentile Summary Graphs

Percentile summary graphs are very similar to cumulative percentage polygons,
but they plot ({\frac{i}{n}},{X_i}), instead of ({X_i}, {\frac{i}{n}}) in a 
scatterplot. Additionally, a percentile summary graph also displays the quantiles
of a distribution, similarly to a boxplot. \

To create a percentile summary graph, two functions are necessary. The first, 
`stat_per_sum()`, creates the scatterplot itsself. Then, `stat_quant()` draws
the quantile lines. Both functions require only one argument, the vector X, as
well as any aesthetics for the quantile lines or dot shape.

## `stat_per_sum()`

```{r}
#example here
```

## `stat_quant()`

```{r}
#example here
```


# Confidence Intervals

```{r}
#example here - may shift this to under the EDF definition
```



